Storm surges in the Strait of Georgia simulated with a regional model

### 2015-08-06 AO_guides/manuscriptAO_final.tex
removed path to figures
Fixed up mistakes from removing color and changing paths to figures

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through {\color{red} the} comparisons of tidal elevation harmonics and storm surge with observations. Important forcing factors contributing to storm surges have been assessed. It is shown that the surge entering the domain from the Pacific Ocean contributes most significantly to the {\color{red} surge amplitude within the Strait of Georgia. Comparisons between simulations with high resolution and low resolution atmospheric forcing further emphasize that remote forcing is the dominant factor in surge amplitudes of this region.} In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generally, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca.
\end{abstract}

@@

The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia {\color{red}\citep{thomson1981oceanography}}. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait (Figure \ref{fig:domain}). At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing {\color{red}\citep{thomson1981oceanography}}.  Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal \citep{thomson1981oceanography}. This system is {\color{red} a} home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modular tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid.  Although these features are not being used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.  NEMO has been successfully used in other coastal modelling studies in the Northeast Atlantic \citep{maraldi2013nemo} and the Mediterranean Sea \citep{brossier2012sensitivity}.}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.pdf}
\caption{ Model domain (highlighted area) including bathymetry, rivers (green circles), and storm surge locations of interest  (red stars). The black rectangle represents the region displayed in Figure \ref{fig:zoom}.}\label{fig:domain}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[scale=0.4]{Figures/bathy_zoom.pdf}
\caption{Model bathymetry and coastlines in the passages between the San Juan and Gulf Islands.}\label{fig:zoom}
\end{figure}

@@

 Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. In support of this assumption, a geostrophic slope calculation with an estuarine velocity of 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca. In the future, this uniform specification of sea level at the western boundary will be relaxed by taking values from the solution of a large-scale forecasting model.}

\begin{table}[h]
\centering 
\tbl{}Average peak surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the five surge events hindcasted in this article. A negative delay indicates that the peak surge occurred later than at Tofino.}
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Port Hardy & 0.491 & -2.67\\
Winter Harbour & 0.572 & 0.333 \\
Tofino  & 0.707 & -- \\
Bamfield & 0.658 & -0.5 \\
Neah Bay & 0.658 & -1.83 \\
Toke Point & 1.17 & -2.83 \\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

@@

%perhaps a table of complex differences at tidal stations similar to Table 1 of Foreman et al (1995)  (such as the one produced by tidetools.calc_diffs_meas_mod) here?
%(would be cool to include the complex differences calculated at the VENUS nodes too)
%perhaps a nice contour map of M2 and K1 amps and phases goes here?
%if it's favourable, we could compare our complex differences to Foreman et al (1995), who got an average of D=3cm for M2 and D=2.5cm for K1

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/tides.pdf}
\caption{Evaluation of model tidal amplitude and phases for the M$_2$ and K$_1$ constituents. On the left, a transect through the stations used for comparison. On the right, modelled M$_2$ (black) and K$_1$ (grey) amplitude (top) and phase (middle) compared with observations in the dashed curves of the same colour. The bottom plot displays the M$_2$ and K$_1$ complex differences. Station names and numbers are listed Appendix \ref{sec:appendix} in Table \ref{tab:comparison}.}
\label{fig:tides}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/feb2006.pdf}
\caption{Comparison of observations and model output for the Feb 4, 2006 storm surge. Left: total water level observation (solid) and corrected model (dashed). Right: observed residuals (solid) and modelled residuals (dashed).}
\label{fig:feb2006}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/feb2006_factors.pdf}
\caption{Comparison of the modelled residual for four simulations in Feb 2006: A simulation with all forcing (solid black), a simulation without local atmospheric forcing (dashed green), a simulation without remote forcing (solid red), and a simulation with only tides and local wind forcing, i.e.\ no remote forcing or local atmospheric pressure (dashed blue). }
\label{fig:factors}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/feb2006_tidesurge.pdf}
\caption{Modelled residuals (solid) compared with a simulation of the surge propagation without tidal forcing (dashed) for the Feb 2006 hindcast. }
\label{fig:tidesurge}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/feb2006_tidesurge_spatial.pdf}
\caption{Spatial dependency of tide-surge interaction in Feb 2006 hindcast. Difference in sea surface height between simulations with tidal forcing and without tidal forcing. Each plot is an average over 4 hours. }
\label{fig:tidesurge_spatial}
\end{figure}

\subsection{Spatial Extent of Surge}\label{sec:spatial}
A major wind storm on Dec 15, 2006 is hindcasted next. Although the storm was very strong (gusts over 120 km hr$^{-1}$) and the observed residual at Point Atkinson reached {\color{red}94 cm}, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. Because of the high winds, this event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006.pdf}
\caption{Comparison of observed and modelled water level and residuals during the Dec 15, 2006 storm surge simulation. }
\label{fig:dec2006}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006_spatial.pdf}
\caption{Propagation of the storm surge on Dec 15, 2006. Modelled residual every hour starting on Dec 15 at 02:30 UTC. Time advances from top left to bottom right. Wind vectors from the atmospheric forcing are overlaid in black.}
\label{fig:spatial}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006_spatial_sshonly.pdf}
\caption{As in Figure \ref{fig:spatial} but for a simulation that does not include atmospheric forcing. }
\label{fig:spatial_sshonly}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/nov2009.pdf}
\caption{Wind and storm surge conditions at Point Atkinson for a strong wind event on Nov 19, 2009. Top: Wind speed from Environment Canada observations \citep{ECClimateArchive} at Point Atkinson (solid) and from the atmospheric model (CGRF) at a grid point near Point Atkinson (dashed). Middle: Sea surface height anomaly at Tofino (solid) and Port Hardy (dashed) used as forcing conditions at the western and northern open boundaries. Bottom: Observed (solid) and modelled (dashed) residuals at Point Atkinson.  }
\label{fig:nov2009}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2012_weather.pdf}
\caption{{\color{red}Comparison of observed and modelled winds during Dec 2012 storm surge simulation at Point Atkinson (top), Sandheads (middle) and NOAA Buoy 46087 (bottom). Observed winds shown in solid black, low resolution CGRF model in dashed black and high resolution model in grey. Observations from Sandheads and Point Atkinson are from the Environment Canada Climate Data archives \citep{ECClimateArchive}.  Observations at the NOAA Buoy were downloaded from: \url{http://www.ndbc.noaa.gov/station_page.php?station=46087}. Wind direction represents the direction that the wind is headed.} }
\label{fig:dec2012_weather}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2012_surge.pdf}
\caption{Comparison of observed (solid black) and modelled residuals for the Dec 2012 storm surge simulation. Modelled residuals from two simulations using a low resolution (dashed) and high resolution (grey) atmospheric products are shown. }
\label{fig:dec2012_surge}
\end{figure}

>>>>

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through the comparisons of tidal elevation harmonics and storm surge with observations. Important forcing factors contributing to storm surges have been assessed. It is shown that the surge entering the domain from the Pacific Ocean contributes most significantly to the surge amplitude within the Strait of Georgia. Comparisons between simulations with high resolution and low resolution atmospheric forcing further emphasize that remote forcing is the dominant factor in surge amplitudes of this region. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas during a major wind storm on Dec 15, 2006. Generally, surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca.
\end{abstract}

@@

The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia \citep{thomson1981oceanography}. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait (Figure \ref{fig:domain}). At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing \citep{thomson1981oceanography}.  Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal \citep{thomson1981oceanography}. This system is a home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modular tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid.  Although these features are not being used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.  NEMO has been successfully used in other coastal modelling studies in the Northeast Atlantic \citep{maraldi2013nemo} and the Mediterranean Sea \citep{brossier2012sensitivity}.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{bathy.pdf}
\caption{ Model domain (highlighted area) including bathymetry, rivers (green circles), and storm surge locations of interest  (red stars). The black rectangle represents the region displayed in Figure \ref{fig:zoom}.}\label{fig:domain}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[scale=0.4]{bathy_zoom.pdf}
\caption{Model bathymetry and coastlines in the passages between the San Juan and Gulf Islands.}\label{fig:zoom}
\end{figure}

@@

 Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. In support of this assumption, a geostrophic slope calculation with an estuarine velocity of 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca. In the future, this uniform specification of sea level at the western boundary will be relaxed by taking values from the solution of a large-scale forecasting model.

\begin{table}[h]
\centering 
\tbl{Average peak surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the five surge events hindcasted in this article. A negative delay indicates that the peak surge occurred later than at Tofino.}
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Port Hardy & 0.491 & -2.67\\
Winter Harbour & 0.572 & 0.333 \\
Tofino  & 0.707 & -- \\
Bamfield & 0.658 & -0.5 \\
Neah Bay & 0.658 & -1.83 \\
Toke Point & 1.17 & -2.83 \\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{tides.pdf}
\caption{Evaluation of model tidal amplitude and phases for the M$_2$ and K$_1$ constituents. On the left, a transect through the stations used for comparison. On the right, modelled M$_2$ (black) and K$_1$ (grey) amplitude (top) and phase (middle) compared with observations in the dashed curves of the same colour. The bottom plot displays the M$_2$ and K$_1$ complex differences. Station names and numbers are listed Appendix \ref{sec:appendix} in Table \ref{tab:comparison}.}
\label{fig:tides}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{feb2006.pdf}
\caption{Comparison of observations and model output for the Feb 4, 2006 storm surge. Left: total water level observation (solid) and corrected model (dashed). Right: observed residuals (solid) and modelled residuals (dashed).}
\label{fig:feb2006}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{feb2006_factors.pdf}
\caption{Comparison of the modelled residual for four simulations in Feb 2006: A simulation with all forcing (solid black), a simulation without local atmospheric forcing (dashed green), a simulation without remote forcing (solid red), and a simulation with only tides and local wind forcing, i.e.\ no remote forcing or local atmospheric pressure (dashed blue). }
\label{fig:factors}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{feb2006_tidesurge.pdf}
\caption{Modelled residuals (solid) compared with a simulation of the surge propagation without tidal forcing (dashed) for the Feb 2006 hindcast. }
\label{fig:tidesurge}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.6]{feb2006_tidesurge_spatial.pdf}
\caption{Spatial dependency of tide-surge interaction in Feb 2006 hindcast. Difference in sea surface height between simulations with tidal forcing and without tidal forcing. Each plot is an average over 4 hours. }
\label{fig:tidesurge_spatial}
\end{figure}

\subsection{Spatial Extent of Surge}\label{sec:spatial}
A major wind storm on Dec 15, 2006 is hindcasted next. Although the storm was very strong (gusts over 120 km hr$^{-1}$) and the observed residual at Point Atkinson reached 94 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. Because of the high winds, this event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{dec2006.pdf}
\caption{Comparison of observed and modelled water level and residuals during the Dec 15, 2006 storm surge simulation. }
\label{fig:dec2006}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{dec2006_spatial.pdf}
\caption{Propagation of the storm surge on Dec 15, 2006. Modelled residual every hour starting on Dec 15 at 02:30 UTC. Time advances from top left to bottom right. Wind vectors from the atmospheric forcing are overlaid in black.}
\label{fig:spatial}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{dec2006_spatial_sshonly.pdf}
\caption{As in Figure \ref{fig:spatial} but for a simulation that does not include atmospheric forcing. }
\label{fig:spatial_sshonly}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{nov2009.pdf}
\caption{Wind and storm surge conditions at Point Atkinson for a strong wind event on Nov 19, 2009. Top: Wind speed from Environment Canada observations \citep{ECClimateArchive} at Point Atkinson (solid) and from the atmospheric model (CGRF) at a grid point near Point Atkinson (dashed). Middle: Sea surface height anomaly at Tofino (solid) and Port Hardy (dashed) used as forcing conditions at the western and northern open boundaries. Bottom: Observed (solid) and modelled (dashed) residuals at Point Atkinson.  }
\label{fig:nov2009}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{dec2012_weather.pdf}
\caption{Comparison of observed and modelled winds during Dec 2012 storm surge simulation at Point Atkinson (top), Sandheads (middle) and NOAA Buoy 46087 (bottom). Observed winds shown in solid black, low resolution CGRF model in dashed black and high resolution model in grey. Observations from Sandheads and Point Atkinson are from the Environment Canada Climate Data archives \citep{ECClimateArchive}.  Observations at the NOAA Buoy were downloaded from: \url{http://www.ndbc.noaa.gov/station_page.php?station=46087}. Wind direction represents the direction that the wind is headed.} 
\label{fig:dec2012_weather}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{dec2012_surge.pdf}
\caption{Comparison of observed (solid black) and modelled residuals for the Dec 2012 storm surge simulation. Modelled residuals from two simulations using a low resolution (dashed) and high resolution (grey) atmospheric products are shown. }
\label{fig:dec2012_surge}
\end{figure}
